Seminar 5
========================================================

### Libraries

```{r}
library(lattice)
library(car)
```

### Reusable functions

```{r}
prepareData <- function(x) {
  pDatDes <- cbind(t(prDat[x,]),prDes)
  pDat <- with(pDatDes,data.frame(sidChar, sidNum, 
                  devStage = factor(devStage,levels=as.character(unique(devStage))),
                  gType = factor(gType,levels=as.character(unique(gType))),
                  gExp = c(unlist(pDatDes[,grepl(".*at",colnames(pDatDes))])),
                  gene = factor(rep(grep(".*at",colnames(pDatDes),value=TRUE), each = nrow(pDatDes))),stringsAsFactors=FALSE))
  row.names(pDat) <- seq(1:nrow(pDat))
  pDat
}
```

```{r}
makeStripplot <- function(data, group = data$gType, type = c('p','a')) {
  stripplot(gExp ~ devStage | gene, data,
          group = group, jitter.data = TRUE,
          auto.key = TRUE, type = type, grid = TRUE)
}
```

### Warm-up exercises

Load the data as usual
```{r}
prDes <- read.table("../GSE4051_design.tsv", header = TRUE, as.is = TRUE)
prDat <- read.table("../GSE4051_data.tsv")
```

Prepare and plot data for probeset "1456341_a_at"
```{r}
testDat <- prepareData("1456341_a_at")
str(testDat)
makeStripplot(testDat)
```

Conduct t-test for probeset "1456341_a_at" at developmental stage P2 vs. 4 weeks post-natal
```{r}
t.test(x=subset(testDat,devStage=="P2",select=c("gExp")),y=subset(testDat,devStage=="4_weeks",select=c("gExp")))
```

Linear model for one categorical covariate (probeset "1438786_a_at")
```{r}
testDat2 <- prepareData("1438786_a_at")
makeStripplot(testDat2)
mfit <- lm(formula = gExp ~ devStage, data = testDat2, subset = gType == "wt")
```

Use this model to compare P2 and P10
```{r}
contMat <- c(0,1,0,-1,0)
diff <- contMat %*% coef(mfit)
estSe <- t(contMat) %*% vcov(mfit) %*% contMat
testStat <- diff/estSe
2 * pt(abs(testStat), df = df.residual(mfit), lower.tail = FALSE)
```

Two categorical covariates: devStage and gType (probeset "1448690_at")
```{r}
makeStripplot(oDat <- prepareData("1448690_at"))
oFitBig <- lm(formula = gExp ~ devStage * gType, data = oDat)
summary(oFitBig)$coef
oFitSmall <- lm(formula = gExp ~ devStage + gType, data = oDat)
summary(oFitSmall)$coef
anova(oFitSmall,oFitBig)
```

Repeating the above for probeset 1429225_at
```{r}
makeStripplot(oDat <- prepareData("1429225_at"))
oFitBig <- lm(formula = gExp ~ devStage * gType, data = oDat)
summary(oFitBig)$coef
oFitSmall <- lm(formula = gExp ~ devStage + gType, data = oDat)
summary(oFitSmall)$coef
anova(oFitSmall,oFitBig)
```

### Take home exercise 1

Now, let's do this for lots of genes
```{r}
set.seed(340)
samp <- sample(1:nrow(prDat), size = 9, replace = FALSE)
luckyGenes <- rownames(prDat)[samp]
nineGeneDat <- prepareData(luckyGenes)
makeStripplot(nineGeneDat)
```

1428334_at looks interesting - could there be an interaction? Let's do some tests...
```{r}
oFitBig <- with(nineGeneDat,by(nineGeneDat, gene, function(x) lm(formula = gExp ~ devStage * gType, data = x)))
oFitSmall <- with(nineGeneDat,by(nineGeneDat, gene, function(x) lm(formula = gExp ~ devStage + gType, data = x)))
coefsBig <- lapply(oFitBig, function(x) summary(x)$coef)
coefsSmall <- lapply(oFitSmall, function(x) summary(x)$coef)
nums <- seq(1:9)
lapply(nums, function(x) anova(oFitSmall[[x]],oFitBig[[x]]))
```

Seems that 3 and 8 (1423641_s_at and 1448504_a_at) have the most evidence for an interaction effect. Not what I would have expected from looking at the plots.

### Take home exercise 2

Let's keep working with our set of nine genes. Let's transform the devStage column into a numeric type column called age.
```{r}
nineGeneDat$age <- recode(nineGeneDat$devStage, "'E16'=-2; 'P2'=2; 'P6'=6; 'P10'=10; '4_weeks'=28",as.factor.result = FALSE)
oFitAge <- with(nineGeneDat,by(nineGeneDat, gene, function(x) lm(formula = gExp ~ age, data = x)))
oFitAgeQuad <- with(nineGeneDat,by(nineGeneDat, gene, function(x) lm(formula = gExp ~ age + I(age^2), data = x)))
coefsAge <- lapply(oFitAge, function(x) summary(x)$coef)
coefsQuad <- lapply(oFitAgeQuad, function(x) summary(x)$coef)
lapply(nums, function(x) anova(oFitAge[[x]],oFitAgeQuad[[x]]))
```

Looks like the quadratic term might be improving the model significantly in case 3, and maybe 1 and 2 depending one what threshold you set. This seems a bit strange given the plot:
```{r}
makeStripplot(nineGeneDat,group=NULL)
```


### Take home exercise 3

Now let's try this again after removing 4 weeks.
```{r}
nineGeneDatFilt <- nineGeneDat[nineGeneDat$devStage != "4_weeks",]
oFitAgeFilt <- with(nineGeneDatFilt,by(nineGeneDatFilt, gene, function(x) lm(formula = gExp ~ age, data = x)))
oFitAgeQuadFilt <- with(nineGeneDatFilt,by(nineGeneDatFilt, gene, function(x) lm(formula = gExp ~ age + I(age^2), data = x)))
coefsAgeFilt <- lapply(oFitAgeFilt, function(x) summary(x)$coef)
coefsQuadFilt <- lapply(oFitAgeQuadFilt, function(x) summary(x)$coef)
lapply(nums, function(x) anova(oFitAgeFilt[[x]],oFitAgeQuadFilt[[x]]))
```

Now the quadratic model looks promising for a few more of the genes.

### Take home exercise 4

Now let's add regression lines to the strip plots.
```{r}
makeStripplot(nineGeneDat,NULL,c('p','r'))
```

