Seminar 6
========================================================

### Libraries

```{r}
library(lattice)
library(limma)
```

### Reusable functions

```{r}
prepareData <- function(x) {
  pDatDes <- cbind(t(prDat[x,]),prDes)
  pDat <- with(pDatDes,data.frame(sidChar, sidNum, 
                  devStage = factor(devStage,levels=as.character(unique(devStage))),
                  gType = factor(gType,levels=as.character(unique(gType))),
                  gExp = c(unlist(pDatDes[,grepl(".*at",colnames(pDatDes))])),
                  gene = factor(rep(grep(".*at",colnames(pDatDes),value=TRUE), each = nrow(pDatDes))),stringsAsFactors=FALSE))
  row.names(pDat) <- seq(1:nrow(pDat))
  pDat
}
```

```{r}
makeStripplot <- function(data, group = data$gType, type = c('p','a')) {
  stripplot(gExp ~ devStage | gene, data,
          group = group, jitter.data = TRUE,
          auto.key = TRUE, type = type, grid = TRUE)
}
```

Stolen from Sohrab...
```{r}
niceScatter <- function(x.var, y.var, x.label, y.label) {
    lims <- c(-20, 15)
    xyplot(y.var ~ x.var, asp = 1, panel = function(x, y, ...) {
        panel.smoothScatter(x, y, ...)
        panel.abline(0, 1, col = "orange")
    }, nbin = 150, ylim = lims, xlim = lims, xlab = x.label, ylab = y.label)
}
```

### Fit a linear model

Load the data as usual.
```{r}
prDes <- readRDS("../GSE4051_design.rds")
prDat <- read.table("../GSE4051_data.tsv")
```

Fit a linear model to the wildtype data.
```{r}
wtDes <- subset(prDes,gType=="wt")
wtDat <- subset(prDat,select=prDes$gType=="wt")
wtDesMat <- model.matrix(~devStage,wtDes)
wtFit <- lmFit(wtDat,wtDesMat)
wtEbFit <- eBayes(wtFit)
```

Find differentially expressed genes.
```{r}
topTable(wtEbFit)
dsHits <- topTable(wtEbFit, coef = grep("devStage", colnames(coef(wtEbFit))))
```

Make plots for three of these hits.
```{r}
pDat <- prepareData(c("1451617_at","1425222_x_at","1422929_s_at"))
wtpDat <- subset(pDat,gType=="wt")
makeStripplot(wtpDat,group=NULL)
```
These plots highlight the differential expression of these genes across dev stages.

This seems to support the hypothesis that devstage has an effect on its own.

Compare to lm.
```{r}
# Get lm results for 1451617_at
wtFitlm <- lm(gExp ~ devStage, wtpDat[1:35,])

summary(wtFitlm)$coef
dsHits["1451617_at",]
```

It's hard to compare these two because lm assigns a p-value to each coef whereas limma assigns a p-value to the probe across all values of devstage. However, they seem to be in the same ballpark.

### topTable exercises

How many probes have Benjamini-Hochberg ("BH") adjusted p-values for the F test conducted above that are less than 1e-05?
```{r}
BHtab <- topTable(wtEbFit,coef = grep("devStage", colnames(coef(wtEbFit))),number=Inf,adjust.method="BH",p.value = 1e-05)
nrow(BHtab)
```

What is the 63rd hit on this list?
```{r}
BHtab[63,c("F","adj.P.Val","devStageP6")]
```

Scatterplot the t statistics for the test that the P2 effect is zero against that for P10.
```{r}
P2tab <- topTable(wtEbFit,coef = grep("devStageP2", colnames(coef(wtEbFit))),number=Inf,sort.by="none")
P10tab <- topTable(wtEbFit,coef = grep("devStageP10", colnames(coef(wtEbFit))),number=Inf,sort.by="none")
niceScatter(P2tab$t,P10tab$t,"P2 t-stat","P10 t-stat")
```

Create a densityplot of the associated adjusted p-values.
```{r}
densityplot(~ P2tab$adj.P.Val + P10tab$adj.P.Val,auto.key=TRUE)
```

If you require a BH adjusted p-value less than 1e-03, how many hits do you get for P2? How many for P10? How much overlap is there?
```{r}
addmargins(table(P2tab$adj.P.Val < 1e-03, P10tab$adj.P.Val < 1e-03, dnn = c("P2", "P10")))
```

Now just focus on the P10 effect. Create a scatterplot matrix of raw p-values, BH adjusted p-values, and BY p-values.
```{r}
P10tab_BY <- topTable(wtEbFit,coef = grep("devStageP10", colnames(coef(wtEbFit))),number=Inf,adjust.method="BY",sort="none")
splomDat <- data.frame(raw=P10tab$P.Value,BH=P10tab$adj.P.Val,BY=P10tab_BY$adj.P.Val)
splom(splomDat)
```
The BH values are almost linearly related to the raw values, although they ascend somewhat faster.

### Perform inference for some contrasts.

Set up contrasts.
```{r}
cont.matrix <- makeContrasts(P10VsP6 = devStageP10 - devStageP6, fourweeksVsP10 = devStage4_weeks - devStageP10, levels = wtDesMat)
wtFitCont <- contrasts.fit(wtFit, cont.matrix)
wtEbFitCont <- eBayes(wtFitCont)
contTab <- topTable(wtEbFitCont)
contTab
```

Plot the data from the top 4 hits.
```{r}
pDat <- prepareData(rownames(contTab)[1:4])
wtpDat <- subset(pDat,gType=="wt")
makeStripplot(wtpDat,group=NULL)
```

Let's use decideTests() to adjust the p-values for both contrasts globally, i.e. all together and then threshhold them at a cutoff of 1e-04.
```{r}
cutoff <- 1e-04
wtResCont <- decideTests(wtEbFitCont, p.value = cutoff, method = "global")
summary(wtResCont)
```

Let's try to pull out various hits and plot their data.
```{r}
hits1 <- rownames(prDat)[which(wtResCont[, "P10VsP6"] < 0)]
hits2 <- rownames(prDat)[which(wtResCont[, "fourweeksVsP10"] < 0)]
hits3 <- rownames(prDat)[which(wtResCont[, "fourweeksVsP10"] > 0)]

pDat_hits1 <- prepareData(hits1)
wtpDat_hits1 <- subset(pDat_hits1,gType=="wt")
makeStripplot(wtpDat_hits1,group=NULL)

pDat_hits2 <- prepareData(hits2[1:4])
wtpDat_hits2 <- subset(pDat_hits2,gType=="wt")
makeStripplot(wtpDat_hits2,group=NULL)

pDat_hits3 <- prepareData(hits3[1:4])
wtpDat_hits3 <- subset(pDat_hits3,gType=="wt")
makeStripplot(wtpDat_hits3,group=NULL)

intersect(hits1,hits2)
intersect(hits1,hits3)
intersect(hits2,hits3)
```

Repeat with less stringent cut-off.
```{r}
cutoff <- 0.01
nhits <- 8
wtResCont <- decideTests(wtEbFitCont, p.value = cutoff, method = "global")
summary(wtResCont)
vennDiagram(wtResCont)

hits1 <- rownames(prDat)[which(wtResCont[, "P10VsP6"] < 0)]
hits2 <- rownames(prDat)[which(wtResCont[, "fourweeksVsP10"] < 0)]
hits3 <- rownames(prDat)[which(wtResCont[, "P10VsP6"] > 0)]
hits4 <- rownames(prDat)[which(wtResCont[, "fourweeksVsP10"] > 0)]
hits5 <- rownames(prDat)[which(wtResCont[, "P10VsP6"] != 0 & wtResCont[, "fourweeksVsP10"] != 0)]
hits6 <- rownames(prDat)[which(wtResCont[, "P10VsP6"] > 0 & wtResCont[, "fourweeksVsP10"] < 0)]

pDat_hits1 <- prepareData(hits1[1:nhits])
wtpDat_hits1 <- subset(pDat_hits1,gType=="wt")
makeStripplot(wtpDat_hits1,group=NULL)

pDat_hits2 <- prepareData(hits2[1:nhits])
wtpDat_hits2 <- subset(pDat_hits2,gType=="wt")
makeStripplot(wtpDat_hits2,group=NULL)

pDat_hits3 <- prepareData(hits3[1:nhits])
wtpDat_hits3 <- subset(pDat_hits3,gType=="wt")
makeStripplot(wtpDat_hits3,group=NULL)

pDat_hits4 <- prepareData(hits4[1:nhits])
wtpDat_hits4 <- subset(pDat_hits4,gType=="wt")
makeStripplot(wtpDat_hits4,group=NULL)

pDat_hits5 <- prepareData(hits5)
wtpDat_hits5 <- subset(pDat_hits5,gType=="wt")
makeStripplot(wtpDat_hits5,group=NULL)

pDat_hits6 <- prepareData(hits6)
wtpDat_hits6 <- subset(pDat_hits6,gType=="wt")
makeStripplot(wtpDat_hits6,group=NULL)
```

### Take-home exercise: See if you can find one or more probes that have some expression changes up to P6 and then hold steady all the way to 4_weeks.

Set up contrasts.
```{r}
cont.matrix <- makeContrasts(P2VsE16 = devStageP2 - Intercept, P6VsP2 = devStageP6 - devStageP2, P10VsP6 = devStageP10 - devStageP6, fourweeksVsP10 = devStage4_weeks - devStageP10, levels = wtDesMat)
wtFitCont <- contrasts.fit(wtFit, cont.matrix)
wtEbFitCont <- eBayes(wtFitCont)
contTab <- topTable(wtEbFitCont)
contTab
```

Try with the more stringent cut-off, since I reckon a lot of probes will meet the criterea (more change expected at beginning, as demonstrated in Venn diagram below).
```{r}
cutoff <- 1e-04
wtResCont <- decideTests(wtEbFitCont, p.value = cutoff, method = "global")
summary(wtResCont)
vennDiagram(wtResCont)
```
Looks like most of the change is going on between E16 and P2 (this makes sense).

Find probes that meet the conditions - changes at both points before P6 but not after P6.
```{r}
hits <- rownames(prDat)[which(wtResCont[, "P2VsE16"] != 0 & wtResCont[, "P6VsP2"] != 0 & wtResCont[, "P10VsP6"] == 0 & wtResCont[, "fourweeksVsP10"] == 0)]

pDat_hits <- prepareData(hits[1:nhits])
wtpDat_hits <- subset(pDat_hits,gType=="wt")
makeStripplot(wtpDat_hits,group=NULL)
```